---
title: "Paired Merged Analysis - CART T0 vs T12 (patient S1)"
author: "K. Charitidis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6
)
```

## Paths

```{r paths}
project_dir <- Sys.getenv("COSMX_PROJECT_DIR",
                          unset = normalizePath("../..", mustWork = TRUE))
scripts_dir <- file.path(project_dir, "Scripts")

source(file.path(scripts_dir, "Merged_Scripts", "Merged_Functions.R"))
```

```{r libs}
suppressPackageStartupMessages({
  library(Giotto); library(GiottoClass)
  library(magrittr)
  library(dplyr); library(tibble); library(readr); library(yaml)
  library(ggplot2)
  library(scry)
  library(scPearsonPCA)
  library(harmony)
  library(BiocNeighbors)
  library(uwot)
  library(igraph)
})
```

```{r config}
cfg <- yaml::read_yaml(file.path(scripts_dir, "Parameters", "config.yaml"))
output_root <- cfg$paths$output_dir %||% file.path(project_dir, "Output")
```

## Pair selection

```{r pair_samples}
pair_ids <- c("CART_T0_S1", "CART_T12_S1")

samples <- readr::read_csv(
  file.path(scripts_dir, "Parameters", "sample_sheet.csv"),
  show_col_types = FALSE
) %>%
  dplyr::filter(sample_id %in% pair_ids) %>%
  dplyr::mutate(output_dir = file.path(output_root, paste0("Sample_", sample_id))) %>%
  dplyr::arrange(match(sample_id, pair_ids))

stopifnot(nrow(samples) == length(pair_ids),
          all(dir.exists(samples$output_dir)))
samples %>%
  dplyr::select(sample_id, patient_id, treatment, timepoint, slide_id, output_dir)
```

## Load raw CosMx data per sample

Mirrors 01_Load_data.R: readr+gzfile (no arrow) -> createGiottoObject -> addCellMetadata.

```{r load_raw_per_sample}
raw_root <- file.path(project_dir, "Data", "Raw_data")
stopifnot(dir.exists(raw_root))

xy_cols <- list(
  x = c("CenterX_global_px", "x_global_px", "CenterX_local_px"),
  y = c("CenterY_global_px", "y_global_px", "CenterY_local_px")
)

per_sample_gobjs <- lapply(seq_len(nrow(samples)), function(i) {
  s <- samples[i, ]
  message("Loading ", s$sample_id, " (FOV ", s$fov_min, "-", s$fov_max, ")")

  sample_dir <- file.path(raw_root, s$directory_name)
  expr_file  <- list.files(sample_dir, "_exprMat_file\\.csv(\\.gz)?$", full.names = TRUE)
  meta_file  <- list.files(sample_dir, "_metadata_file\\.csv(\\.gz)?$", full.names = TRUE)
  stopifnot(length(expr_file) == 1, length(meta_file) == 1)

  read_csv_gz <- function(p) as.data.frame(
    readr::read_csv(if (grepl("\\.gz$", p)) gzfile(p) else p,
                    show_col_types = FALSE)
  )
  expr <- read_csv_gz(expr_file)
  meta <- read_csv_gz(meta_file)

  if (!is.na(s$fov_min) && !is.na(s$fov_max)) {
    fov_range <- seq(s$fov_min, s$fov_max)
    expr <- expr[expr$fov %in% fov_range, , drop = FALSE]
    meta <- meta[meta$fov %in% fov_range, , drop = FALSE]
  }

  expr$gid <- paste(expr$fov, expr$cell_ID, sep = "_")
  meta$gid <- paste(meta$fov, meta$cell_ID, sep = "_")
  common   <- intersect(expr$gid, meta$gid)
  expr     <- expr[match(common, expr$gid), , drop = FALSE]
  meta     <- meta[match(common, meta$gid), , drop = FALSE]

  gene_cols <- setdiff(names(expr), c("fov", "cell_ID", "gid"))
  counts    <- Matrix::Matrix(t(as.matrix(expr[, gene_cols])), sparse = TRUE)
  colnames(counts) <- expr$gid
  rownames(counts) <- gene_cols

  x_col <- xy_cols$x[xy_cols$x %in% names(meta)][1]
  y_col <- xy_cols$y[xy_cols$y %in% names(meta)][1]
  spatial_locs <- data.frame(
    cell_ID = meta$gid,
    sdimx   = as.numeric(meta[[x_col]]),
    sdimy   = as.numeric(meta[[y_col]])
  )

  meta$cell_ID   <- meta$gid
  meta$sample_id <- s$sample_id
  g <- Giotto::createGiottoObject(expression = counts, spatial_locs = spatial_locs)
  Giotto::addCellMetadata(g, new_metadata = meta,
                          by_column = TRUE, column_cell_ID = "cell_ID")
})
names(per_sample_gobjs) <- samples$sample_id

tibble::tibble(
  sample_id = names(per_sample_gobjs),
  n_cells   = vapply(per_sample_gobjs,
                     function(g) nrow(GiottoClass::pDataDT(g)), integer(1))
)
```

## Per-sample cell QC

Per sample: features-floor -> min_count + cell_max_genes -> WTx +/-2.5 SD log10 outlier -> SplitRatioToLocal border.

```{r qc_per_sample}
qc <- cfg$parameters$qc

# Compute all flags on the raw object, capture for plotting, apply final pass once.
qc_results <- lapply(seq_along(per_sample_gobjs), function(i) {
  g  <- per_sample_gobjs[[i]]
  id <- names(per_sample_gobjs)[i]

  g      <- Giotto::addStatistics(g, expression_values = "raw")
  meta   <- as.data.frame(GiottoClass::pDataDT(g))
  raw    <- Giotto::getExpression(g, values = "raw", output = "matrix")
  coords <- as.data.frame(GiottoClass::getSpatialLocations(g, output = "data.table"))
  ord    <- match(meta$cell_ID, coords$cell_ID)
  meta$sdimx <- coords$sdimx[ord]
  meta$sdimy <- coords$sdimy[ord]

  is_neg <- grepl("^(NegPrb|SystemControl)", rownames(raw))
  meta$negprb_ratio <- if (any(is_neg)) {
    nr <- Matrix::colSums(raw[is_neg, , drop = FALSE]) / Matrix::colSums(raw)
    unname(nr[match(meta$cell_ID, names(nr))])
  } else NA_real_

  meta$flag_low_features  <- meta$nr_feats   <  qc$cell_min_genes
  meta$flag_low_counts    <- meta$total_expr <  qc$min_count
  meta$flag_high_features <- meta$nr_feats   >  qc$cell_max_genes

  # WTx +/-2.5 SD log10 outlier on the cells passing the floor + window (matches sequential WTx logic).
  prefilt <- !meta$flag_low_features & !meta$flag_low_counts & !meta$flag_high_features
  for (col in c("total_expr", "nr_feats")) {
    flag_col <- if (col == "total_expr") "flag_count_outlier" else "flag_feat_outlier"
    meta[[flag_col]] <- FALSE
    if (sum(prefilt) > 1L) {
      lv  <- log10(meta[[col]][prefilt])
      lo  <- mean(lv) - 2.5 * sd(lv)
      hi  <- mean(lv) + 2.5 * sd(lv)
      lvf <- log10(meta[[col]])
      meta[[flag_col]] <- (lvf < lo | lvf > hi)
    }
  }

  # SplitRatioToLocal border flag (Area / mean(Area | fov) < 0.5).
  area_col <- intersect(c("Area", "Area.um2", "cell_area"), names(meta))[1]
  meta$flag_border <- FALSE
  if (!is.na(area_col)) {
    ratio <- meta[[area_col]] / ave(meta[[area_col]], meta$fov,
                                     FUN = function(x) mean(x, na.rm = TRUE))
    meta$flag_border <- ratio < 0.5
  }

  meta$qc_pass <- !meta$flag_low_features & !meta$flag_low_counts &
                  !meta$flag_high_features & !meta$flag_count_outlier &
                  !meta$flag_feat_outlier  & !meta$flag_border

  n_before <- nrow(meta)
  n_after  <- sum(meta$qc_pass)
  message(sprintf("  %s: %d -> %d cells (%.1f%% retained)",
                  id, n_before, n_after, 100 * n_after / n_before))

  list(
    gobj  = Giotto::subsetGiotto(g, cell_ids = meta$cell_ID[meta$qc_pass]),
    flags = meta
  )
})

per_sample_gobjs    <- lapply(qc_results, `[[`, "gobj")
qc_flags_per_sample <- lapply(qc_results, `[[`, "flags")
names(per_sample_gobjs)    <- samples$sample_id
names(qc_flags_per_sample) <- samples$sample_id

tibble::tibble(
  sample_id = names(per_sample_gobjs),
  n_cells   = vapply(per_sample_gobjs,
                     function(g) nrow(GiottoClass::pDataDT(g)), integer(1))
)
```

## QC visualisations

Pre-filter views per sample using `qc_flags_per_sample`; thresholds drawn as dashed lines.

```{r qc_viz, fig.width=11, fig.height=5}
qc_all <- dplyr::bind_rows(qc_flags_per_sample)
flag_cols   <- c("flag_low_features", "flag_low_counts", "flag_high_features",
                 "flag_count_outlier", "flag_feat_outlier", "flag_border")
flag_labels <- c("Low features", "Low counts", "High features",
                 "Count outlier", "Feature outlier", "Border (low SplitRatio)")
pass_pal    <- c(`TRUE` = "dodgerblue", `FALSE` = "darkorange")

print(ggplot(qc_all, aes(total_expr)) +
  geom_density(fill = "grey80", colour = NA) +
  geom_vline(xintercept = qc$min_count, linetype = "dashed") +
  scale_x_log10() +
  facet_wrap(~ sample_id, scales = "free_y") +
  labs(x = "nCount_RNA (log10)", y = "Density",
       title = "Transcripts per cell"))

print(ggplot(qc_all, aes(nr_feats)) +
  geom_density(fill = "grey80", colour = NA) +
  geom_vline(xintercept = c(qc$cell_min_genes, qc$cell_max_genes),
             linetype = "dashed") +
  scale_x_log10() +
  facet_wrap(~ sample_id, scales = "free_y") +
  labs(x = "nFeature_RNA (log10)", y = "Density",
       title = "Features per cell"))

print(ggplot(qc_all, aes(nr_feats, total_expr, colour = qc_pass)) +
  geom_point(size = 0.3, alpha = 0.5) +
  geom_vline(xintercept = c(qc$cell_min_genes, qc$cell_max_genes),
             linetype = "dashed") +
  geom_hline(yintercept = qc$min_count, linetype = "dashed") +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = pass_pal) +
  facet_wrap(~ sample_id) +
  labs(x = "nFeature_RNA (log10)", y = "nCount_RNA (log10)",
       colour = "QC pass", title = "Joint count/feature distribution"))

print(ggplot(qc_all, aes(sdimx, sdimy, colour = qc_pass)) +
  geom_point(size = 0.2, alpha = 0.5) +
  scale_colour_manual(values = pass_pal) +
  coord_fixed() +
  facet_grid(qc_pass ~ sample_id, switch = "y") +
  labs(title = "Spatial arrangement of flagged cells"))

qc_neg <- qc_all %>% dplyr::filter(!is.na(negprb_ratio))
if (nrow(qc_neg) > 0L) {
  print(ggplot(qc_neg,
               aes(sdimx, sdimy, colour = log2(pmax(negprb_ratio, 1e-4)))) +
    geom_point(size = 0.2, alpha = 0.6) +
    scale_colour_viridis_c(option = "magma") +
    coord_fixed() +
    facet_wrap(~ sample_id) +
    labs(colour = "log2(NegPrb / Total)",
         title = "Negative-probe ratio per cell"))
}

upset_input           <- as.data.frame(qc_all[, flag_cols])
upset_input[]         <- lapply(upset_input, as.integer)
colnames(upset_input) <- flag_labels
UpSetR::upset(upset_input, sets = flag_labels,
              order.by = "freq", nintersects = 10,
              mainbar.y.label = "Cells with flag combination",
              sets.x.label    = "Cells flagged")
```

## QC diagnostic - negative-probe ratio

Print-only sanity check. Healthy CosMx WTx data -> median negprb ratio < 1-2%.

```{r qc_diagnostics_negprb}
diag_negprb <- lapply(seq_along(per_sample_gobjs), function(i) {
  g   <- per_sample_gobjs[[i]]
  raw <- Giotto::getExpression(g, values = "raw", output = "matrix")
  is_neg <- grepl("^(NegPrb|SystemControl)", rownames(raw))
  if (!any(is_neg)) return(NULL)
  ratio <- Matrix::colSums(raw[is_neg, , drop = FALSE]) / Matrix::colSums(raw)
  tibble::tibble(
    sample_id    = names(per_sample_gobjs)[i],
    n_neg_probes = sum(is_neg),
    median_ratio = median(ratio, na.rm = TRUE),
    p95_ratio    = unname(quantile(ratio, 0.95, na.rm = TRUE))
  )
})
dplyr::bind_rows(diag_negprb)
```

## FOV QC (Nanostring scratch-space)

Per-FOV bias + signal-loss flagging via `runFOVQC`; "flag-and-monitor", no auto-filter.

```{r fov_qc}
fovqc_cache <- tools::R_user_dir("CosMxNotebook", which = "cache")
dir.create(fovqc_cache, recursive = TRUE, showWarnings = FALSE)

utils_path    <- file.path(fovqc_cache, "FOV_QC_utils.R")
barcodes_path <- file.path(fovqc_cache, "barcodes_by_panel.RDS")
if (!file.exists(utils_path)) {
  download.file(
    "https://raw.githubusercontent.com/Nanostring-Biostats/CosMx-Analysis-Scratch-Space/Main/_code/FOV%20QC/FOV%20QC%20utils.R",
    utils_path
  )
}
if (!file.exists(barcodes_path)) {
  download.file(
    "https://github.com/Nanostring-Biostats/CosMx-Analysis-Scratch-Space/raw/Main/_code/FOV%20QC/barcodes_by_panel.RDS",
    barcodes_path, mode = "wb"
  )
}
source(utils_path)
barcodemap <- readRDS(barcodes_path)$Hs_6k

fov_qc <- lapply(seq_along(per_sample_gobjs), function(i) {
  g      <- per_sample_gobjs[[i]]
  raw    <- Giotto::getExpression(g, values = "raw", output = "matrix")
  meta   <- GiottoClass::pDataDT(g)
  coords <- GiottoClass::getSpatialLocations(g, output = "data.table")
  ord    <- match(colnames(raw), meta$cell_ID)
  res <- runFOVQC(
    counts               = Matrix::t(raw),
    xy                   = as.matrix(coords[ord, .(sdimx, sdimy)]),
    fov                  = meta$fov[ord],
    barcodemap           = barcodemap,
    max_prop_loss        = 0.4,
    max_totalcounts_loss = 0.3
  )
  tibble::tibble(
    sample_id    = names(per_sample_gobjs)[i],
    n_flagged    = length(res$flaggedfovs),
    flagged_fovs = paste(res$flaggedfovs, collapse = ",")
  )
})
dplyr::bind_rows(fov_qc)
```

## Merge

```{r merge_paths}
merged_label <- paste0("Paired_", paste(samples$sample_id, collapse = "_"))
merged_dir   <- file.path(output_root, "Merged_Notebook", merged_label)
dir.create(merged_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r merge}
merged_gobj <- Giotto::joinGiottoObjects(
  gobject_list  = per_sample_gobjs,
  gobject_names = samples$sample_id,
  join_method   = cfg$merged$join_method %||% "shift",
  x_padding     = cfg$merged$x_padding   %||% 5000
)
merged_gobj
```

## Cohort-level gene QC

Single cohort threshold; toggle off via `qc$apply_gene_filter: false` to follow WTx guide.

```{r qc_post_merge_genes}
if (isTRUE(qc$apply_gene_filter %||% TRUE)) {
  merged_gobj <- Giotto::filterGiotto(
    gobject                = merged_gobj,
    expression_threshold   = 1,
    feat_det_in_min_cells  = qc$gene_min_cells,
    min_det_feats_per_cell = 0,
    expression_values      = "raw"
  )
}
merged_gobj
```

```{r merged_summary}
merged_meta <- tibble::as_tibble(GiottoClass::pDataDT(merged_gobj))
raw_counts  <- Giotto::getExpression(merged_gobj, values = "raw", output = "matrix")
cat("Merged cells:", nrow(merged_meta),
    "| Genes:", nrow(raw_counts), "\n")
merged_meta %>% dplyr::count(sample_id, name = "n_cells")
```

## HVG - binomial deviance (WTx-style)

```{r hvg_pr}
n_hvgs <- .wtx(cfg, "n_hvgs", 4000L)

dev <- scry::devianceFeatureSelection(raw_counts, fam = "binomial")
hvg_ids <- names(sort(dev, decreasing = TRUE))[seq_len(min(n_hvgs, length(dev)))]

cat("HVGs selected:", length(hvg_ids), "/", nrow(raw_counts), "\n")
head(hvg_ids, 10)
```

## PCA - sparse quasi-Poisson (scPearsonPCA)

```{r pca_pr}
hvg_counts <- raw_counts[hvg_ids, , drop = FALSE]
total_per_cell <- Matrix::colSums(raw_counts)
total_per_gene <- Matrix::rowSums(raw_counts)
target_freq    <- (total_per_gene / sum(total_per_gene))[hvg_ids]

pcaobj <- scPearsonPCA::sparse_quasipoisson_pca_seurat(
  x           = hvg_counts,
  totalcounts = total_per_cell,
  grate       = target_freq,
  scale.max   = .wtx(cfg, "pca_scale_max",  10),
  do.scale    = .wtx(cfg, "pca_do_scale",   TRUE),
  do.center   = .wtx(cfg, "pca_do_center",  TRUE),
  ncores      = .wtx(cfg, "pca_ncores",     5L)
)
pca_mat <- pcaobj$reduction.data@cell.embeddings
n_pcs   <- min(.wtx(cfg, "n_pcs", 40L), ncol(pca_mat))
pca_mat <- pca_mat[, seq_len(n_pcs), drop = FALSE]
dim(pca_mat)
```

## Harmony batch correction

```{r harmony}
batch_col <- .wtx(cfg, "harmony_batch_column", "sample_id")
batch_vec <- merged_meta[[batch_col]][match(rownames(pca_mat), merged_meta$cell_ID)]

harmony_pcs <- harmony::HarmonyMatrix(
  data_mat  = pca_mat,
  meta_data = data.frame(batch = batch_vec),
  vars_use  = "batch",
  do_pca    = FALSE,
  verbose   = FALSE
)
dim(harmony_pcs)
```

## kNN graph (cosine)

```{r knn}
k <- .wtx(cfg, "knn_k", 15L)

knn <- BiocNeighbors::findKNN(
  harmony_pcs,
  k       = k,
  BNPARAM = BiocNeighbors::AnnoyParam(distance = "Cosine")
)
str(knn, max.level = 1)
```

## UMAP

```{r umap}
umap_xy <- uwot::umap(
  harmony_pcs,
  n_neighbors = k,
  min_dist    = .wtx(cfg, "umap_min_dist", 0.02),
  spread      = .wtx(cfg, "umap_spread",   2),
  metric      = .wtx(cfg, "knn_metric",    "cosine"),
  init        = "spectral"
)
colnames(umap_xy) <- c("UMAP_1", "UMAP_2")
merged_meta <- merged_meta %>%
  dplyr::mutate(UMAP_1 = umap_xy[, 1], UMAP_2 = umap_xy[, 2])
```

## Leiden clustering

Jaccard SNN graph (bluster) + modularity-objective Leiden; matches the scanpy/Seurat default.

```{r leiden}
snn <- bluster::makeSNNGraph(
  harmony_pcs,
  k    = .wtx(cfg, "knn_k", 15L),
  type = "jaccard"
)

leiden_part <- igraph::cluster_leiden(
  snn,
  objective_function   = "modularity",
  resolution_parameter = .wtx(cfg, "leiden_resolution",   0.5),
  n_iterations         = .wtx(cfg, "leiden_n_iterations", 2L),
  weights              = igraph::E(snn)$weight
)
merged_meta$leiden_clust <- factor(igraph::membership(leiden_part))

merged_gobj <- Giotto::addCellMetadata(
  merged_gobj,
  new_metadata    = data.frame(
    cell_ID      = merged_meta$cell_ID,
    leiden_clust = merged_meta$leiden_clust,
    UMAP_1       = merged_meta$UMAP_1,
    UMAP_2       = merged_meta$UMAP_2,
    stringsAsFactors = FALSE
  ),
  by_column       = TRUE,
  column_cell_ID  = "cell_ID"
)
```

## Cluster summary + UMAPs

```{r cluster_summary}
merged_meta %>%
  dplyr::count(sample_id, leiden_clust, name = "n_cells") %>%
  tidyr::pivot_wider(names_from = sample_id, values_from = n_cells, values_fill = 0)
```

```{r umap_plots, fig.width=11, fig.height=5}
p_clust <- ggplot(merged_meta, aes(UMAP_1, UMAP_2, colour = leiden_clust)) +
  geom_point(size = 0.4, alpha = 0.8) +
  theme_bw() +
  labs(title = "Leiden clusters", colour = NULL)

p_sample <- ggplot(merged_meta, aes(UMAP_1, UMAP_2, colour = sample_id)) +
  geom_point(size = 0.4, alpha = 0.8) +
  theme_bw() +
  labs(title = "Sample of origin", colour = NULL)

cowplot::plot_grid(p_clust, p_sample, nrow = 1)
```

---

> Next: cell-type annotation on the merged object. HieraType vs marker-gene
> Wilcoxon vs InSituType - open question.
