---
title: "Paired Merged Analysis — CART T0 vs T12 (patient S1)"
author: "K. Charitidis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 6
)
```

## Overview

Stand-alone notebook for a *paired* CART biopsy comparison (T0 vs T12, same
patient). Built outside `CosMx_pipeline.R` so the merge / batch / DE / pathway
choices can be tuned by hand for the paired design. Reusable helpers will
accumulate in `Merged_Functions.R`; one-off logic stays in this file.

The pair below is the working test pair (`CART_T0_S1`, `CART_T12_S1`).
Re-point the `pair_samples` chunk to use a different pair.

## Setup

```{r paths}
project_dir <- normalizePath(file.path(here::here(), ".."), mustWork = FALSE)
if (!dir.exists(project_dir)) {
  project_dir <- normalizePath("..", mustWork = TRUE)
}

merged_fn <- file.path(project_dir, "Merged_Scripts", "Merged_Functions.R")
source(merged_fn, local = FALSE)

# Pull in the production load + merge helpers so this notebook stays in
# step with the pipeline's join / re-normalization behaviour.
source(file.path(project_dir, "CosMx_pipeline.R"),
       local = FALSE, chdir = TRUE)
source(file.path(project_dir, "Helper_Scripts", "Merge_Batch_Correction.R"),
       local = FALSE, chdir = TRUE)
```

```{r libs}
suppressPackageStartupMessages({
  library(Giotto)
  library(GiottoClass)
  library(dplyr)
  library(tibble)
  library(readr)
  library(yaml)
})
```

```{r config}
config_path <- file.path(project_dir, "Parameters", "config.yaml")
cfg <- yaml::read_yaml(config_path)

# Resolve the per-sample output root the same way the pipeline does. The
# pipeline writes per-sample artifacts to <output_dir>/Sample_<sample_id>/
# (final layout, see Run_Scripts/Run_Giotto_Pipeline.sh).
output_root <- cfg$paths$output_dir %||% file.path(project_dir, "Output")
stopifnot(dir.exists(output_root))
output_root
```

## Pair selection

```{r pair_samples}
sample_sheet_path <- file.path(project_dir, "Parameters", "sample_sheet.csv")
sample_sheet <- readr::read_csv(sample_sheet_path, show_col_types = FALSE)

pair_ids <- c("CART_T0_S1", "CART_T12_S1")

samples <- sample_sheet |>
  dplyr::filter(sample_id %in% pair_ids) |>
  dplyr::mutate(
    output_dir = file.path(output_root, paste0("Sample_", sample_id))
  ) |>
  dplyr::arrange(match(sample_id, pair_ids))

stopifnot(nrow(samples) == length(pair_ids))
all(dir.exists(samples$output_dir)) ||
  stop("Per-sample output dirs missing: ",
       paste(samples$output_dir[!dir.exists(samples$output_dir)],
             collapse = ", "))

samples |>
  dplyr::select(sample_id, patient_id, treatment, timepoint, slide_id,
                fov_min, fov_max, output_dir)
```

## Load the per-sample Giotto checkpoints

Loads the `07_annotate` checkpoint for each sample in the pair (default
`merged.input_step` from config; override if needed). Polygon centroids are
stripped before the join — `joinGiottoObjects()` regenerates them, and stale
slot contents trigger a centroid/polygon count mismatch.

```{r load_checkpoints}
input_step <- cfg$merged$input_step %||% "07_annotate"

per_sample_gobjs <- lapply(seq_len(nrow(samples)), function(i) {
  message("Loading ", samples$sample_id[i], " (", input_step, ")")
  gobj <- load_sample_checkpoint(samples[i, , drop = FALSE], input_step)
  .strip_polygon_centroids(gobj)
})
names(per_sample_gobjs) <- samples$sample_id

# Quick-look summary for each loaded object.
loaded_summary <- tibble::tibble(
  sample_id   = names(per_sample_gobjs),
  n_cells     = vapply(per_sample_gobjs,
                       function(g) nrow(.giotto_pdata_dt(g)),
                       integer(1)),
  metadata_cols = vapply(per_sample_gobjs,
                         function(g) ncol(.giotto_pdata_dt(g)),
                         integer(1))
)
loaded_summary
```

## Merge the two objects

Uses the production `merge_giotto_samples()` (from
`Helper_Scripts/Merge_Batch_Correction.R`) so the shift-join, x-padding, and
re-normalization behaviour match the pipeline.

The output dir for this notebook is held separate from the pipeline's
`Output/Merged/` runs to avoid clobbering pipeline artifacts. It mirrors the
naming scheme so cross-referencing remains easy.

```{r merge_paths}
merged_label <- paste0("Paired_", paste(samples$sample_id, collapse = "_"))
merged_output_dir_nb <- file.path(output_root, "Merged_Notebook", merged_label)
dir.create(merged_output_dir_nb, recursive = TRUE, showWarnings = FALSE)
merged_output_dir_nb
```

```{r merge}
merged_gobj <- merge_giotto_samples(
  gobject_list  = per_sample_gobjs,
  sample_table  = samples,
  output_dir    = merged_output_dir_nb,
  join_method   = cfg$merged$join_method %||% "shift",
  x_padding     = cfg$merged$x_padding %||% 5000,
  renormalize   = cfg$merged$renormalize %||% TRUE,
  scalefactor   = cfg$merged$normalization_scalefactor %||% NULL,
  log_transform = cfg$merged$normalization_log_transform %||% TRUE,
  save_object   = TRUE
)
```

## Inspect the merged object

```{r merged_summary}
merged_meta <- tibble::as_tibble(.giotto_pdata_dt(merged_gobj))

cat("Merged cells: ", nrow(merged_meta), "\n",
    "Merged genes: ", nrow(Giotto::getExpression(merged_gobj,
                                                  values = "raw",
                                                  output = "matrix")),
    "\n", sep = "")

merged_meta |>
  dplyr::count(sample_id, name = "n_cells")
```

```{r merged_meta_preview}
merged_meta |>
  dplyr::select(dplyr::any_of(c("cell_ID", "sample_id", "patient_id",
                                "treatment", "timepoint", "leiden_clust",
                                "cell_type", "nr_feats", "total_expr"))) |>
  head(10)
```

---

> Next: decide what the *paired* analysis layer should look like — batch
> correction strategy, paired DE backend, paired CCI handling. Update this
> notebook chunk-by-chunk; promote helpers to `Merged_Functions.R` only
> when they get reused across chunks (and confirm before promoting).
