---
title: "Paired Merged Analysis — CART T0 vs T12 (patient S1)"
author: "K. Charitidis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6
)
```

## Paths

```{r paths}
project_dir <- Sys.getenv("COSMX_PROJECT_DIR",
                          unset = normalizePath("../..", mustWork = TRUE))
scripts_dir <- file.path(project_dir, "Scripts")

source(file.path(scripts_dir, "Merged_Scripts", "Merged_Functions.R"))
source(file.path(scripts_dir, "CosMx_pipeline.R"), chdir = TRUE)
```

```{r libs}
suppressPackageStartupMessages({
  library(Giotto); library(GiottoClass)
  library(magrittr)
  library(dplyr); library(tibble); library(readr); library(yaml)
  library(ggplot2)
  library(scry)
  library(scPearsonPCA)
  library(harmony)
  library(BiocNeighbors)
  library(uwot)
  library(igraph)
})
```

```{r config}
cfg <- yaml::read_yaml(file.path(scripts_dir, "Parameters", "config.yaml"))
output_root <- cfg$paths$output_dir %||% file.path(project_dir, "Output")
```

## Pair selection

```{r pair_samples}
pair_ids <- c("CART_T0_S1", "CART_T12_S1")

samples <- readr::read_csv(
  file.path(scripts_dir, "Parameters", "sample_sheet.csv"),
  show_col_types = FALSE
) %>%
  dplyr::filter(sample_id %in% pair_ids) %>%
  dplyr::mutate(output_dir = file.path(output_root, paste0("Sample_", sample_id))) %>%
  dplyr::arrange(match(sample_id, pair_ids))

stopifnot(nrow(samples) == length(pair_ids),
          all(dir.exists(samples$output_dir)))
samples %>%
  dplyr::select(sample_id, patient_id, treatment, timepoint, slide_id, output_dir)
```

## Load post-QC checkpoints

Loads each sample's `Giotto_Object_Filtered` (step 02). Norm / dim reduction /
clustering / annotation are redone jointly on the merged object below.

```{r load_checkpoints}
input_step <- "02_qc"

per_sample_gobjs <- lapply(seq_len(nrow(samples)), function(i) {
  message("Loading ", samples$sample_id[i], " @ ", input_step)
  .strip_polygon_centroids(
    load_sample_checkpoint(samples[i, , drop = FALSE], input_step)
  )
})
names(per_sample_gobjs) <- samples$sample_id

tibble::tibble(
  sample_id = names(per_sample_gobjs),
  n_cells   = vapply(per_sample_gobjs,
                     function(g) nrow(.giotto_pdata_dt(g)), integer(1))
)
```

## Merge + renormalize

```{r merge_paths}
merged_label <- paste0("Paired_", paste(samples$sample_id, collapse = "_"))
merged_dir   <- file.path(output_root, "Merged_Notebook", merged_label)
dir.create(merged_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r merge}
merged_gobj <- Giotto::joinGiottoObjects(
  gobject_list  = per_sample_gobjs,
  gobject_names = samples$sample_id,
  join_method   = cfg$merged$join_method %||% "shift",
  x_padding     = cfg$merged$x_padding   %||% 5000
)
merged_gobj
```

```{r merged_summary}
merged_meta <- tibble::as_tibble(.giotto_pdata_dt(merged_gobj))
raw_counts  <- .gobj_raw_counts(merged_gobj)
cat("Merged cells:", nrow(merged_meta),
    "| Genes:", nrow(raw_counts), "\n")
merged_meta %>% dplyr::count(sample_id, name = "n_cells")
```

## HVG — binomial deviance (WTx-style)

```{r hvg_pr}
n_hvgs <- .wtx(cfg, "n_hvgs", 4000L)

dev <- scry::devianceFeatureSelection(raw_counts, fam = "binomial")
hvg_ids <- names(sort(dev, decreasing = TRUE))[seq_len(min(n_hvgs, length(dev)))]

cat("HVGs selected:", length(hvg_ids), "/", nrow(raw_counts), "\n")
head(hvg_ids, 10)
```

## PCA — sparse quasi-Poisson (scPearsonPCA)

```{r pca_pr}
hvg_counts <- raw_counts[hvg_ids, , drop = FALSE]
total_per_cell <- Matrix::colSums(raw_counts)
total_per_gene <- Matrix::rowSums(raw_counts)
target_freq    <- (total_per_gene / sum(total_per_gene))[hvg_ids]

pcaobj <- scPearsonPCA::sparse_quasipoisson_pca_seurat(
  x           = hvg_counts,
  totalcounts = total_per_cell,
  grate       = target_freq,
  scale.max   = .wtx(cfg, "pca_scale_max",  10),
  do.scale    = .wtx(cfg, "pca_do_scale",   TRUE),
  do.center   = .wtx(cfg, "pca_do_center",  TRUE),
  ncores      = .wtx(cfg, "pca_ncores",     5L)
)
pca_mat <- pcaobj$reduction.data@cell.embeddings
n_pcs   <- min(.wtx(cfg, "n_pcs", 40L), ncol(pca_mat))
pca_mat <- pca_mat[, seq_len(n_pcs), drop = FALSE]
dim(pca_mat)
```

## Harmony batch correction

```{r harmony}
batch_col <- .wtx(cfg, "harmony_batch_column", "sample_id")
batch_vec <- merged_meta[[batch_col]][match(rownames(pca_mat), merged_meta$cell_ID)]

harmony_pcs <- harmony::HarmonyMatrix(
  data_mat  = pca_mat,
  meta_data = data.frame(batch = batch_vec),
  vars_use  = "batch",
  do_pca    = FALSE,
  verbose   = FALSE
)
dim(harmony_pcs)
```

## kNN graph (cosine)

```{r knn}
k <- .wtx(cfg, "knn_k", 15L)

knn <- BiocNeighbors::findKNN(
  harmony_pcs,
  k       = k,
  BNPARAM = BiocNeighbors::AnnoyParam(distance = "Cosine")
)
str(knn, max.level = 1)
```

## UMAP

```{r umap}
umap_xy <- uwot::umap(
  harmony_pcs,
  n_neighbors = k,
  min_dist    = .wtx(cfg, "umap_min_dist", 0.02),
  spread      = .wtx(cfg, "umap_spread",   2),
  metric      = .wtx(cfg, "knn_metric",    "cosine"),
  init        = "spectral"
)
colnames(umap_xy) <- c("UMAP_1", "UMAP_2")
merged_meta <- merged_meta %>%
  dplyr::mutate(UMAP_1 = umap_xy[, 1], UMAP_2 = umap_xy[, 2])
```

## Leiden clustering

```{r leiden}
edge_list <- do.call(rbind, lapply(seq_len(nrow(knn$index)), function(i) {
  cbind(i, knn$index[i, ])
}))
g <- igraph::graph_from_edgelist(edge_list, directed = FALSE) %>%
  igraph::simplify()

leiden_part <- igraph::cluster_leiden(
  g,
  resolution_parameter = .wtx(cfg, "leiden_resolution",   0.5),
  n_iterations         = .wtx(cfg, "leiden_n_iterations", 2L)
)
merged_meta$leiden_clust <- factor(igraph::membership(leiden_part))

merged_gobj <- .gobj_add_meta(
  merged_gobj,
  data.frame(cell_ID      = merged_meta$cell_ID,
             leiden_clust = merged_meta$leiden_clust,
             UMAP_1       = merged_meta$UMAP_1,
             UMAP_2       = merged_meta$UMAP_2,
             stringsAsFactors = FALSE)
)
```

## Cluster summary + UMAPs

```{r cluster_summary}
merged_meta %>%
  dplyr::count(sample_id, leiden_clust, name = "n_cells") %>%
  tidyr::pivot_wider(names_from = sample_id, values_from = n_cells, values_fill = 0)
```

```{r umap_plots, fig.width=11, fig.height=5}
p_clust <- ggplot(merged_meta, aes(UMAP_1, UMAP_2, colour = leiden_clust)) +
  geom_point(size = 0.4, alpha = 0.8) +
  theme_bw() +
  labs(title = "Leiden clusters", colour = NULL)

p_sample <- ggplot(merged_meta, aes(UMAP_1, UMAP_2, colour = sample_id)) +
  geom_point(size = 0.4, alpha = 0.8) +
  theme_bw() +
  labs(title = "Sample of origin", colour = NULL)

cowplot::plot_grid(p_clust, p_sample, nrow = 1)
```

---

> Next: cell-type annotation on the merged object. HieraType vs marker-gene
> Wilcoxon vs InSituType — open question.
