---
title: "Merged-run Summary: `r params$run_label`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    df_print: paged
params:
  merged_output_dir: ""
  run_label: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      results = "asis")
out_dir <- params$merged_output_dir
read_csv_safe <- function(p) {
  if (!file.exists(p)) return(NULL)
  tryCatch(utils::read.csv(p, stringsAsFactors = FALSE),
           error = function(e) NULL)
}
read_text_safe <- function(p) {
  if (!file.exists(p)) return(NULL)
  paste(readLines(p, warn = FALSE), collapse = "\n")
}
```

> **Warning: Slide × arm confound, read first.**
> All CART biopsies sit on slide 1; CONV samples sit on slides 3, 4, 31, 32;
> Control on slide 2. Treatment arm therefore cannot be cleanly separated
> from slide-level technical variation. **Cross-arm comparisons at any
> single timepoint (e.g. CART vs CONV at T0) are confounded with slide and
> should be read as suggestive only.** The clean comparisons are
> within-arm pre/post (paired by patient) and the `arm × timepoint`
> interaction on per-patient deltas.
>
> Secondary note: `sample_id` Harmony also un-pairs Conv slides 3+4 and
> 31+32 which were technically run together (accepted tradeoff per user
> direction).

# 1. Run information

```{r run-info}
manifest <- read_csv_safe(file.path(out_dir, "10_Merged",
                                    "merged_sample_manifest.csv"))
if (!is.null(manifest)) {
  cat("Samples in this merged run:\n\n")
  knitr::kable(manifest)
} else {
  cat("_No `merged_sample_manifest.csv` found._\n")
}
```

# 2. Post-Harmony diagnostics

```{r diagnostics}
diag_md <- read_text_safe(file.path(out_dir, "10_Merged",
                                    "diagnostics", "summary.md"))
if (!is.null(diag_md)) {
  cat(diag_md)
} else {
  cat("_No diagnostics output found at `10_Merged/diagnostics/summary.md`. ",
      "Phase 3 may not have run._\n")
}
```

# 3. Cell-type composition (`propeller`)

```{r composition}
comp_dir <- file.path(out_dir, "15_Composition")
if (dir.exists(comp_dir)) {
  files <- list.files(comp_dir, pattern = "\\.csv$", full.names = TRUE)
  if (length(files) == 0) {
    cat("_`15_Composition/` exists but contains no CSVs._\n")
  } else {
    for (f in files) {
      cat("\n## ", tools::file_path_sans_ext(basename(f)), "\n\n", sep = "")
      df <- read_csv_safe(f)
      if (!is.null(df) && nrow(df) > 0) {
        df <- df[order(df$padj), , drop = FALSE]
        knitr::kable(utils::head(df, 20),
                     caption = "Top 20 cell types by adjusted p")
      }
    }
  }
} else {
  cat("_No composition tests output (`15_Composition/`). Phase 6 may not have run._\n")
}
```

# 4. BANKSY niches

```{r banksy}
banksy_csv <- file.path(out_dir, "14_BANKSY", "niche_summary.csv")
df <- read_csv_safe(banksy_csv)
if (!is.null(df)) {
  knitr::kable(df, caption = "Cells per niche per sample")
} else {
  cat("_No BANKSY output (`14_BANKSY/niche_summary.csv`). Phase 7 may not have run._\n")
}
```

# 5. Pseudobulk DE, top hits per comparison

```{r de-top, results = "asis"}
pw_dir <- file.path(out_dir, "13_Pathway_Analysis")
if (dir.exists(pw_dir)) {
  comps <- list.dirs(pw_dir, recursive = FALSE)
  comps <- comps[!grepl("(^|/)(_|leading_edge|bcell_focus)", comps)]
  for (cdir in comps) {
    cat("\n## ", basename(cdir), "\n\n", sep = "")
    de_files <- list.files(cdir, pattern = "_de_ranks\\.csv$",
                           recursive = TRUE, full.names = TRUE)
    for (f in utils::head(de_files, 5)) {
      df <- read_csv_safe(f)
      if (is.null(df) || nrow(df) == 0) next
      cat("\n### ", basename(dirname(f)), "\n\n", sep = "")
      df <- df[order(df$padj), , drop = FALSE]
      print(knitr::kable(utils::head(df[, c("gene", "logFC", "pval", "padj",
                                             "engine", "design")], 15)))
    }
  }
} else {
  cat("_No pathway output (`13_Pathway_Analysis/`)._\n")
}
```

# 6. Paired GSEA-score sub-step

```{r paired-gsea, results = "asis"}
pgs <- list.files(pw_dir, pattern = "paired_gsea_scores\\.csv$",
                  recursive = TRUE, full.names = TRUE)
if (length(pgs) == 0) {
  cat("_Paired GSEA-score sub-step did not produce output. ",
      "Either disabled (`cfg$pathway$paired_gsea_scores$enabled`) or per-sample step 13 hasn't been run yet._\n")
} else {
  for (f in pgs) {
    df <- read_csv_safe(f)
    if (is.null(df) || nrow(df) == 0) next
    relpath <- sub(paste0(out_dir, "/?"), "", f)
    cat("\n## ", relpath, "\n\n", sep = "")
    df <- df[order(df$padj), , drop = FALSE]
    print(knitr::kable(utils::head(df, 20),
                        caption = "Top 20 paired-pathway hits"))
  }
}
```

# 7. Paired CCI (LIANA -> paired Wilcoxon)

```{r cci-paired, results = "asis"}
cci_dir <- file.path(out_dir, "16_CCI_Paired")
if (dir.exists(cci_dir)) {
  files <- list.files(cci_dir, pattern = "_paired_lrs\\.csv$", full.names = TRUE)
  for (f in files) {
    df <- read_csv_safe(f)
    if (is.null(df) || nrow(df) == 0) next
    cat("\n## ", tools::file_path_sans_ext(basename(f)), "\n\n", sep = "")
    df <- df[order(df$padj), , drop = FALSE]
    print(knitr::kable(utils::head(df, 20),
                        caption = "Top 20 paired LR pairs"))
  }
} else {
  cat("_No paired CCI output (`16_CCI_Paired/`). Phase 10 may not have run._\n")
}
```

# 8. smiDE paired-blocking audit

```{r smide-audit, results = "asis"}
sde_dir <- file.path(out_dir, "12_Spatial_Differential_Expression")
csvs <- list.files(sde_dir, pattern = "_results\\.csv$|_spatial_de\\.csv$",
                   recursive = TRUE, full.names = TRUE)
if (length(csvs) > 0) {
  audit_rows <- list()
  for (f in csvs) {
    df <- read_csv_safe(f)
    if (is.null(df) || !"paired_blocking" %in% names(df)) next
    pb <- unique(df$paired_blocking)
    audit_rows[[length(audit_rows) + 1L]] <- data.frame(
      file = sub(paste0(out_dir, "/?"), "", f),
      paired_blocking = paste(pb, collapse = " / "),
      n_rows = nrow(df), stringsAsFactors = FALSE)
  }
  if (length(audit_rows) > 0) {
    audit <- do.call(rbind, audit_rows)
    knitr::kable(audit,
                 caption = paste("Per-stratum smiDE blocking, strata flagged",
                                 "`skipped_single_patient` are unblocked and",
                                 "should not be over-interpreted as paired."))
  } else {
    cat("_smiDE results found but no `paired_blocking` column._\n")
  }
} else {
  cat("_No smiDE output found in `12_Spatial_Differential_Expression/`._\n")
}
```

---

_Generated by `Helper_Scripts/Render_Merged_Summary.R` (Stage 1, Phase 13)._
