---
title: "CosMx Analysis — Results Summary"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
params:
  project_dir:       !r Sys.getenv("COSMX_PROJECT_DIR", unset = NA_character_)
  output_root:       "Output"
  sample_sheet:      "Parameters/sample_sheet.csv"
  config_path:       "Parameters/config.yaml"
  merged_tables_dir: "Output/merged_figures/Merged_Tables"
  top_n_markers:     25
  top_n_liana:       25
  top_n_ligands:     10
  smide_celltypes:   !r c("B cell", "CD4 T cell", "CD8 T cell")
  bcell_label:       "B cell"
  fdr_threshold:     0.05
  fail_soft:         true
---

```{css, echo=FALSE}
.section-box { background:#f8f9fa; border-left:4px solid #2c7bb6; padding:10px 16px; margin:12px 0; border-radius:3px; }
.stat-inline  { font-weight:bold; color:#2c7bb6; }
.warn-box     { background:#fff8e1; border-left:4px solid #f0ad4e; padding:8px 14px; margin:10px 0; border-radius:3px; }
table.kable-table { font-size:0.88em; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(purrr)
  library(stringr); library(yaml); library(jsonlite); library(knitr); library(tibble)
})
has_kex <- requireNamespace("kableExtra", quietly=TRUE)
if (has_kex) library(kableExtra)

`%||%` <- function(a, b) if (is.null(a) || length(a)==0 || identical(a,"")) b else a

# ---- path resolution -------------------------------------------------------
resolve_project_dir <- function(p) {
  if (!is.na(p) && nzchar(p) && dir.exists(p)) return(normalizePath(p, mustWork=TRUE))
  cands <- c(normalizePath("..", mustWork=FALSE), normalizePath("../..", mustWork=FALSE), getwd())
  hit   <- cands[file.exists(file.path(cands, "Parameters", "sample_sheet.csv"))]
  if (!length(hit)) stop("Cannot resolve project_dir. Set COSMX_PROJECT_DIR.")
  normalizePath(hit[1], mustWork=TRUE)
}

project_dir      <- resolve_project_dir(params$project_dir)
config_abs       <- file.path(project_dir, params$config_path)
cfg              <- if (file.exists(config_abs)) yaml::read_yaml(config_abs) else list()

resolve_path <- function(cfg_rel, fallback_rel) {
  if (!is.null(cfg_rel) && nzchar(cfg_rel))
    normalizePath(file.path(dirname(config_abs), cfg_rel), mustWork=FALSE)
  else
    normalizePath(file.path(project_dir, fallback_rel), mustWork=FALSE)
}

output_root_abs    <- resolve_path(cfg$paths$output_dir,  params$output_root)
sample_sheet_abs   <- resolve_path(cfg$paths$sample_sheet, params$sample_sheet)
merged_tables_abs  <- normalizePath(
  file.path(project_dir, params$merged_tables_dir), mustWork=FALSE)
dir.create(merged_tables_abs, recursive=TRUE, showWarnings=FALSE)

# ---- global params ---------------------------------------------------------
top_n_markers   <- cfg$parameters$markers$top_n %||% params$top_n_markers
top_n_liana     <- params$top_n_liana
top_n_ligands   <- params$top_n_ligands
smide_celltypes <- params$smide_celltypes
bcell_label     <- params$bcell_label
fdr_threshold   <- params$fdr_threshold
fail_soft       <- isTRUE(params$fail_soft)

# ---- helpers ---------------------------------------------------------------
read_csv_safe <- function(path) {
  if (!length(path) || is.na(path) || !file.exists(path)) {
    if (!fail_soft) stop("Missing: ", path)
    return(NULL)
  }
  tryCatch(readr::read_csv(path, show_col_types=FALSE, progress=FALSE),
           error=function(e){ message("[err] ", basename(path), ": ", e$message); NULL })
}

find_sample_dir <- function(sid, root) {
  d <- file.path(root, paste0("Sample_", sid))
  if (!dir.exists(d)) { message("[skip] missing: ", d); return(NA_character_) }
  d
}

write_table <- function(df, name) {
  p <- file.path(merged_tables_abs, name)
  readr::write_csv(df, p)
  message("[wrote] ", name, " (", nrow(df), " rows)")
  invisible(p)
}

kt <- function(df, caption=NULL, ...) {
  t <- knitr::kable(df, caption=caption, format.args=list(big.mark=","), ...)
  if (has_kex)
    t <- kableExtra::kable_styling(t, bootstrap_options=c("striped","hover","condensed"),
                                   full_width=FALSE, position="left", font_size=13)
  t
}

pct_fmt <- function(x) paste0(round(x*100, 1), "%")
```

---

# 1. Samples

```{r samples}
sheet <- read_csv_safe(sample_sheet_abs)
stopifnot(!is.null(sheet))

samples_df <- sheet %>%
  filter(treatment %in% c("CART","Conventional"),
         is.na(split_role) | split_role != "composite",
         sample_id != "TEST_5k") %>%
  mutate(sample_dir = vapply(sample_id, find_sample_dir, character(1), root=output_root_abs)) %>%
  select(any_of(c("sample_id","treatment","timepoint","patient_id","pair_id","sample_dir"))) %>%
  arrange(treatment, timepoint, sample_id)

if (nrow(samples_df) != 8)
  warning("Expected 8 samples, got ", nrow(samples_df))
```

<div class="section-box">
**<span class="stat-inline">`r nrow(samples_df)`</span> samples** included:
`r paste(samples_df$sample_id, collapse=" · ")`
</div>

```{r samples_table}
kt(samples_df %>% select(-sample_dir),
   caption="Sample index")
```

---

# 2. Composition

All three clustering levels (Leiden, supervised cell type, BANKSY niches) merged into one table.

```{r composition}
# ---- Leiden ----------------------------------------------------------------
leiden_comp <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    df <- read_csv_safe(file.path(sample_dir, "05_Clustering",
                                  paste0(sample_id, "_cluster_stats.csv")))
    if (is.null(df)) return(NULL)
    df %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint,
                  cluster_type="leiden", label=as.character(leiden_clust)) %>%
      select(sample_id, treatment, timepoint, cluster_type, label, n_cells)
  }) %>%
  group_by(sample_id) %>%
  mutate(pct=n_cells/sum(n_cells)) %>% ungroup()

# ---- Cell type (supervised) ------------------------------------------------
read_selected_profile <- function(json_path) {
  if (!file.exists(json_path)) return(NA_character_)
  j <- tryCatch(jsonlite::read_json(json_path, simplifyVector=FALSE), error=function(e) NULL)
  if (is.null(j)) return(NA_character_)
  pick <- function(x) if (!is.null(x) && length(x)==1L && nzchar(as.character(x))) as.character(x) else NULL
  col <- pick(j$selected_annotation_column) %||% pick(j$selected_col) %||%
         pick(j$default_profile) %||%
         (if (length(j$selected_annotation_columns)>0) pick(j$selected_annotation_columns[[1]]) else NULL)
  if (is.null(col)) return(NA_character_)
  prof <- sub("^celltype_","",col); prof <- sub("_supervised_refined$|_supervised$|_semi$","",prof)
  if (!nzchar(prof)) NA_character_ else prof
}

celltype_comp <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    profile <- read_selected_profile(
      file.path(sample_dir, "07_Annotation", "annotation_selection.json"))
    if (is.na(profile)) return(NULL)
    p_dir <- file.path(sample_dir, "07_Annotation", profile)
    path  <- if (file.exists(file.path(p_dir, paste0(sample_id,"_supervised_summary.csv"))))
      file.path(p_dir, paste0(sample_id,"_supervised_summary.csv")) else
      file.path(p_dir, paste0(sample_id,"_semi_summary.csv"))
    df <- read_csv_safe(path); if (is.null(df)) return(NULL)
    ct_col <- grep("^celltype_", names(df), value=TRUE)[1]
    if (!is.na(ct_col)) names(df)[names(df)==ct_col] <- "celltype"
    df %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint,
                  cluster_type="celltype", label=as.character(celltype)) %>%
      select(sample_id, treatment, timepoint, cluster_type, label, n_cells)
  }) %>%
  group_by(sample_id) %>%
  mutate(pct=n_cells/sum(n_cells)) %>% ungroup()

# ---- BANKSY niches ---------------------------------------------------------
banksy_comp <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    p <- file.path(sample_dir, "14_BANKSY", "BANKSY",
                   paste0(sample_id, "_niche_summary.csv"))
    df <- read_csv_safe(p); if (is.null(df)) return(NULL)
    freq_col <- intersect(c("Freq","n","n_cells","count"), names(df))[1]
    niche_col <- intersect(c("niche_id","niche","label"), names(df))[1]
    if (is.na(freq_col) || is.na(niche_col)) return(NULL)
    df %>% filter(.data[[freq_col]] > 0) %>%
      mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint,
             cluster_type="banksy", label=as.character(.data[[niche_col]]),
             n_cells=as.integer(.data[[freq_col]])) %>%
      select(sample_id, treatment, timepoint, cluster_type, label, n_cells)
  }) %>%
  group_by(sample_id) %>%
  mutate(pct=n_cells/sum(n_cells)) %>% ungroup()

composition_all <- bind_rows(leiden_comp, celltype_comp, banksy_comp) %>%
  mutate(pct_label=pct_fmt(pct)) %>%
  arrange(treatment, timepoint, sample_id, cluster_type, label)

write_table(composition_all, "01_composition_all.csv")
```

<div class="section-box">
Composition table written: **<span class="stat-inline">`r nrow(composition_all)`</span> rows** across
`r n_distinct(composition_all$sample_id)` samples and
`r n_distinct(composition_all$cluster_type)` cluster types
(`r paste(sort(unique(composition_all$cluster_type)), collapse=", ")`).
</div>

```{r composition_display, results="asis"}
# Wide matrix per cluster type for quick reading
for (ctype in c("celltype","leiden","banksy")) {
  sub <- composition_all %>% filter(cluster_type==ctype) %>%
    mutate(val=paste0(n_cells, " (", pct_label, ")")) %>%
    select(label, sample_id, val) %>%
    tidyr::pivot_wider(names_from=sample_id, values_from=val, values_fill="-") %>%
    arrange(label)
  if (nrow(sub)==0) next
  cat("\n\n### ", tools::toTitleCase(ctype), " composition\n\n")
  print(kt(sub, caption=paste(tools::toTitleCase(ctype), "— cells (% of sample)")))
}
```

---

# 3. Differential expression

## 3a. Leiden cluster markers

Significant markers (FDR < `r fdr_threshold`) from one-vs-all scran testing across all samples.

```{r leiden_markers}
leiden_markers <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    df <- read_csv_safe(file.path(sample_dir, "06_Markers", "tables",
                                   paste0(sample_id, "_all_markers.csv")))
    if (is.null(df)) return(NULL)
    df %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint, .before=1)
  })

if (nrow(leiden_markers) > 0) {
  fdr_col <- intersect(c("FDR","padj","p.value","p_val"), names(leiden_markers))[1]
  lfc_col <- intersect(c("summary.logFC","summary_logFC","logFC"), names(leiden_markers))[1]
  gene_col <- intersect(c("gene","feats","feature"), names(leiden_markers))[1]

  leiden_sig <- leiden_markers %>%
    filter(!is.na(.data[[fdr_col]]), .data[[fdr_col]] < fdr_threshold)
  if (!is.na(lfc_col)) leiden_sig <- leiden_sig %>% filter(.data[[lfc_col]] > 0)

  write_table(leiden_sig, "02a_leiden_markers_significant.csv")

  leiden_counts <- leiden_sig %>%
    group_by(sample_id, treatment, timepoint) %>%
    summarise(n_sig_associations=n(), n_sig_genes=n_distinct(.data[[gene_col]]),
              n_clusters=n_distinct(cluster), .groups="drop") %>%
    arrange(treatment, timepoint, sample_id)
} else {
  leiden_sig    <- tibble()
  leiden_counts <- tibble()
}
```

<div class="section-box">
**<span class="stat-inline">`r nrow(leiden_sig)`</span>** significant gene-cluster associations
across **<span class="stat-inline">`r n_distinct(leiden_sig[[gene_col %||% "gene"]])`</span>** unique genes
(FDR < `r fdr_threshold`, logFC > 0).
</div>

```{r leiden_markers_display, results="asis"}
if (nrow(leiden_counts) > 0) {
  print(kt(leiden_counts,
           col.names=c("Sample","Treatment","Timepoint",
                       "Sig. associations","Unique sig. genes","Clusters"),
           caption="Leiden marker summary per sample"))
}

# Top N per sample for HTML preview
if (nrow(leiden_sig) > 0) {
  top_leiden <- leiden_sig %>%
    group_by(sample_id, cluster) %>%
    arrange(.data[[fdr_col]], .by_group=TRUE) %>%
    slice_head(n=top_n_markers) %>% ungroup()
  for (sid in unique(top_leiden$sample_id)) {
    cat("\n\n#### ", sid, "\n\n")
    print(kt(top_leiden %>% filter(sample_id==sid),
             caption=paste("Top", top_n_markers, "markers per cluster —", sid)))
  }
}
```

## 3b. Cell-type annotation markers

Significant markers from one-vs-all scran testing per annotated cell type.

```{r celltype_markers}
celltype_markers <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    df <- read_csv_safe(file.path(sample_dir, "07_Annotation", "tables",
                                   paste0(sample_id, "_annotation_all_markers.csv")))
    if (is.null(df)) return(NULL)
    df %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint, .before=1)
  })

if (nrow(celltype_markers) > 0) {
  fdr_col_ct  <- intersect(c("FDR","padj","p.value","p_val"), names(celltype_markers))[1]
  lfc_col_ct  <- intersect(c("summary.logFC","summary_logFC","logFC"), names(celltype_markers))[1]
  gene_col_ct <- intersect(c("gene","feats","feature"), names(celltype_markers))[1]

  celltype_sig <- celltype_markers %>%
    filter(!is.na(.data[[fdr_col_ct]]), .data[[fdr_col_ct]] < fdr_threshold)
  if (!is.na(lfc_col_ct)) celltype_sig <- celltype_sig %>% filter(.data[[lfc_col_ct]] > 0)

  write_table(celltype_sig, "02b_celltype_markers_significant.csv")

  ct_counts <- celltype_sig %>%
    group_by(sample_id, treatment, timepoint) %>%
    summarise(n_sig_associations=n(), n_sig_genes=n_distinct(.data[[gene_col_ct]]),
              n_celltypes=n_distinct(cluster), .groups="drop") %>%
    arrange(treatment, timepoint, sample_id)

  # B-cell gene spotlight
  bcell_genes <- c("CD19","MS4A1","CD79A","CD79B","IGHG1","IGHG2","IGHM","IGKC",
                   "IGLC1","IGLC2","IGLL1","FCRL4","PRDM1","IRF4","XBP1",
                   "CD38","SDC1","PTPRC","CXCR5","BCL6","AICDA")
  bcell_spotlight <- celltype_markers %>%
    filter(.data[[gene_col_ct]] %in% bcell_genes,
           !is.na(.data[[fdr_col_ct]]), .data[[fdr_col_ct]] < fdr_threshold) %>%
    select(sample_id, treatment, timepoint, cluster,
           gene=all_of(gene_col_ct),
           any_of(c(fdr_col_ct, lfc_col_ct))) %>%
    arrange(sample_id, cluster, .data[[gene_col_ct]])
} else {
  celltype_sig    <- tibble()
  ct_counts       <- tibble()
  bcell_spotlight <- tibble()
}
```

<div class="section-box">
**<span class="stat-inline">`r nrow(celltype_sig)`</span>** significant gene-cell-type associations
across **<span class="stat-inline">`r n_distinct(celltype_sig[[gene_col_ct %||% "gene"]])`</span>** unique genes.
</div>

```{r celltype_markers_display, results="asis"}
if (nrow(ct_counts) > 0) {
  print(kt(ct_counts,
           col.names=c("Sample","Treatment","Timepoint",
                       "Sig. associations","Unique sig. genes","Cell types"),
           caption="Cell-type marker summary per sample"))
}

if (nrow(bcell_spotlight) > 0) {
  cat("\n\n### B-cell / plasma-cell gene spotlight\n\n")
  print(kt(bcell_spotlight, caption="B-cell and plasma-cell genes in cell-type comparisons"))
} else {
  cat('\n\n<div class="warn-box">No canonical B-cell or plasma-cell marker genes were significant in any cell-type comparison.</div>\n\n')
}

if (nrow(celltype_sig) > 0) {
  top_ct <- celltype_sig %>%
    group_by(sample_id, cluster) %>%
    arrange(.data[[fdr_col_ct]], .by_group=TRUE) %>%
    slice_head(n=top_n_markers) %>% ungroup()
  for (sid in unique(top_ct$sample_id)) {
    cat("\n\n#### ", sid, "\n\n")
    print(kt(top_ct %>% filter(sample_id==sid),
             caption=paste("Top", top_n_markers, "cell-type markers —", sid)))
  }
}
```

---

# 4. Spatial DE (smiDE)

Niche-stratified differential expression from RDS fit objects. Cell types:
`r paste(smide_celltypes, collapse=", ")`.

```{r smide}
smide_rds_dir_fn <- function(sample_dir)
  file.path(sample_dir, "12_Spatial_Differential_Expression", "tables")

smide_results_path <- function(sample_dir, sid, celltype) {
  ct_file <- gsub(" ","_",celltype)
  file.path(smide_rds_dir_fn(sample_dir),
            paste0(sid,"_sample_",ct_file,"_smide_spatial_niche_smide_results_raw.rds"))
}
smide_csv_path <- function(sample_dir, sid, celltype) {
  ct_file <- gsub(" ","_",celltype)
  file.path(smide_rds_dir_fn(sample_dir),
            paste0(sid,"_sample_",ct_file,"_smide_spatial_niche_spatial_de.csv"))
}

extract_smide_df <- function(obj) {
  if (is.null(obj))        return(NULL)
  if (is.data.frame(obj))  return(obj)
  if (is.list(obj)) {
    if ("results" %in% names(obj)) return(extract_smide_df(obj$results))
    df_items <- vapply(obj, is.data.frame, logical(1))
    if (any(df_items)) {
      out <- lapply(names(obj)[df_items], function(nm) {
        df <- obj[[nm]]; if (nrow(df)==0L) return(NULL)
        df$result_component <- nm; df
      })
      out <- out[!vapply(out, is.null, logical(1))]
      if (!length(out)) return(NULL)
      dplyr::bind_rows(out)
    }
  }
}

smide_all <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    purrr::map_dfr(smide_celltypes, function(ct) {
      rds_p <- smide_results_path(sample_dir, sample_id, ct)
      csv_p <- smide_csv_path(sample_dir, sample_id, ct)
      df <- if (file.exists(rds_p)) {
        raw <- tryCatch(readRDS(rds_p), error=function(e){ message("[err] ",basename(rds_p)); NULL })
        extract_smide_df(raw)
      } else if (file.exists(csv_p)) {
        read_csv_safe(csv_p)
      } else {
        message("[skip] no smiDE for ", sample_id, " / ", ct); NULL
      }
      if (is.null(df) || nrow(df)==0) return(NULL)
      df %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint,
                    cell_type=ct, .before=1)
    })
  })

if (nrow(smide_all) > 0) {
  smide_sig_col  <- intersect(c("FDR","padj","p.adj","p.value","pval"), names(smide_all))[1]
  smide_fc_col   <- intersect(c("logFC","log2fc","log2FoldChange","estimate"), names(smide_all))[1]
  smide_gene_col <- intersect(c("target","gene","feats"), names(smide_all))[1]
  smide_niche_col <- intersect(c("contrast","term","spatial_niche"), names(smide_all))[1]

  smide_sig <- smide_all %>%
    filter(!is.na(.data[[smide_sig_col]]), .data[[smide_sig_col]] < fdr_threshold)

  write_table(smide_sig, "03_smide_significant.csv")

  smide_counts <- smide_sig %>%
    group_by(sample_id, treatment, timepoint, cell_type) %>%
    summarise(n_sig=n(), n_genes=n_distinct(.data[[smide_gene_col]], na.rm=TRUE),
              .groups="drop") %>%
    arrange(treatment, timepoint, sample_id, cell_type)
} else {
  smide_sig    <- tibble()
  smide_counts <- tibble()
}
```

<div class="section-box">
**<span class="stat-inline">`r nrow(smide_sig)`</span>** significant niche-stratified DE associations
(FDR < `r fdr_threshold`) across `r n_distinct(smide_sig[["cell_type"]] %||% character(0))` cell types.
</div>

```{r smide_display, results="asis"}
if (nrow(smide_counts) > 0) {
  print(kt(smide_counts,
           col.names=c("Sample","Treatment","Timepoint","Cell type","N sig.","N genes"),
           caption="Significant smiDE associations per sample and cell type"))
}

if (nrow(smide_sig) > 0) {
  for (ct in smide_celltypes) {
    sub <- smide_sig %>% filter(cell_type==ct)
    if (nrow(sub)==0) next
    top_s <- if (!is.na(smide_fc_col))
      sub %>% arrange(desc(abs(.data[[smide_fc_col]]))) %>% slice_head(n=50)
    else sub %>% slice_head(n=50)
    cat("\n\n### ", ct, " — top hits\n\n")
    print(kt(top_s, caption=paste(ct, "— top 50 smiDE hits by |logFC|")))
  }
}
```

---

# 5. Cell-cell interactions

## 5a. LIANA

```{r liana}
liana_all <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    df <- read_csv_safe(file.path(sample_dir, "10_CCI_Analysis", "liana",
                                   paste0(sample_id,"_liana_aggregate.csv")))
    if (is.null(df)) return(NULL)
    df %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint, .before=1)
  })

if (nrow(liana_all) > 0) {
  rank_col <- intersect(c("aggregate_rank","liana_score","natmi","sca"), names(liana_all))[1]
  asc      <- identical(rank_col, "aggregate_rank")

  liana_top <- liana_all %>%
    group_by(sample_id) %>%
    arrange(if (asc) .data[[rank_col]] else dplyr::desc(.data[[rank_col]]), .by_group=TRUE) %>%
    slice_head(n=top_n_liana) %>% ungroup()

  # B-cell sender or receiver
  liana_bcell <- liana_all %>%
    filter(if_any(any_of(c("source","sender","sender_label")),
                  ~grepl(bcell_label, .x, fixed=TRUE)) |
           if_any(any_of(c("target","receiver")),
                  ~grepl(bcell_label, .x, fixed=TRUE)))

  write_table(liana_top,   "04a_liana_top.csv")
  write_table(liana_bcell, "04b_liana_bcell.csv")
} else {
  liana_top <- liana_bcell <- tibble()
}
```

<div class="section-box">
LIANA: **<span class="stat-inline">`r nrow(liana_all)`</span>** total LR interactions across all samples.
Top `r top_n_liana` per sample written to `04a_liana_top.csv`.
B-cell interactions (**`r nrow(liana_bcell)`** rows) written to `04b_liana_bcell.csv`.
</div>

```{r liana_display, results="asis"}
if (nrow(liana_top) > 0) {
  for (sid in unique(liana_top$sample_id)) {
    cat("\n\n#### ", sid, "\n\n")
    print(kt(liana_top %>% filter(sample_id==sid),
             caption=paste("Top", top_n_liana, "LIANA interactions —", sid)))
  }
}
if (nrow(liana_bcell) > 0) {
  cat("\n\n### B-cell interactions\n\n")
  print(kt(head(liana_bcell, 60), caption=paste0("LIANA: ",bcell_label," as sender or receiver (head 60)")))
}
```

## 5b. NicheNet

```{r nichenet}
nn_df <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    nn_root  <- file.path(sample_dir, "10_CCI_Analysis", "nichenet")
    cmp_path <- file.path(nn_root, paste0(sample_id,"_nichenet_spatially_retained_comparisons.csv"))
    if (!file.exists(cmp_path))
      cmp_path <- file.path(nn_root, paste0(sample_id,"_nichenet_requested_comparisons.csv"))
    act_path <- file.path(nn_root, paste0(sample_id,"_ligand_activities.csv"))
    cmp <- read_csv_safe(cmp_path)
    act <- read_csv_safe(act_path)
    if (is.null(act)) return(NULL)
    if ("...1" %in% names(act)) act <- act %>% select(-`...1`)
    score_col <- intersect(c("aupr_corrected","aupr","pearson"), names(act))[1]
    if (is.na(score_col)) score_col <- names(act)[2]
    bcell_senders <- if (!is.null(cmp) && "receiver" %in% names(cmp))
      unique(cmp$sender_label[cmp$receiver==bcell_label]) else character(0)
    if (!length(bcell_senders)) {
      message("[skip] no '", bcell_label, "' receiver for ", sample_id); return(NULL)
    }
    top_lig <- act %>%
      arrange(dplyr::desc(.data[[score_col]])) %>% slice_head(n=top_n_ligands)
    tidyr::expand_grid(sender_label=bcell_senders, receiver=bcell_label) %>%
      dplyr::cross_join(top_lig) %>%
      mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint, .before=1)
  })

write_table(if (nrow(nn_df)>0) nn_df else tibble(), "04c_nichenet_ligands.csv")
```

<div class="section-box">
NicheNet: **<span class="stat-inline">`r nrow(nn_df)`</span>** ligand-activity rows for `r bcell_label` as receiver
across `r n_distinct(nn_df[["sample_id"]] %||% character(0))` samples.
</div>

```{r nichenet_display}
if (nrow(nn_df) > 0)
  print(kt(nn_df, caption=paste0("NicheNet top-",top_n_ligands," ligands — ",bcell_label," receiver")))
```

## 5c. MISTy

```{r misty}
read_misty <- function(sample_id, sample_dir, treatment, timepoint, suffix="") {
  p <- file.path(sample_dir, "10_CCI_Analysis", "misty", sample_id, "tables",
                 paste0(sample_id, if (nzchar(suffix)) paste0("_",suffix) else "",
                        "_misty_improvements.csv"))
  df <- read_csv_safe(p); if (is.null(df)) return(NULL)
  if ("...1" %in% names(df)) df <- df %>% select(-`...1`)
  df %>% rename_with(tolower, any_of("View")) %>%
    mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint,
           misty_run=if (nzchar(suffix)) suffix else "hvg", .before=1)
}

read_misty_importances <- function(sample_id, sample_dir, treatment, timepoint, suffix="") {
  imp_dir <- file.path(sample_dir, "10_CCI_Analysis", "misty", sample_id)
  if (!dir.exists(imp_dir)) return(NULL)
  tryCatch({
    res <- mistyR::collect_results(imp_dir)
    imp <- if (!is.null(res$importances.aggregated)) res$importances.aggregated
           else if (!is.null(res$importances)) as.data.frame(res$importances)
           else NULL
    if (is.null(imp)) return(NULL)
    imp %>% mutate(sample_id=sample_id, treatment=treatment, timepoint=timepoint,
                   misty_run=if (nzchar(suffix)) suffix else "hvg", .before=1)
  }, error=function(e){ message("[skip] mistyR::collect_results failed for ",sample_id); NULL })
}

misty_imp_df <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    bind_rows(
      read_misty(sample_id, sample_dir, treatment, timepoint, ""),
      read_misty(sample_id, sample_dir, treatment, timepoint, "bcell")
    )
  })

misty_importances <- samples_df %>% filter(!is.na(sample_dir)) %>%
  pmap_dfr(function(sample_id, sample_dir, treatment, timepoint, ...) {
    read_misty_importances(sample_id, sample_dir, treatment, timepoint)
  })

write_table(if (nrow(misty_imp_df)>0) misty_imp_df else tibble(), "04d_misty_improvements.csv")
if (nrow(misty_importances)>0)
  write_table(misty_importances, "04e_misty_importances.csv")
```

<div class="section-box">
MISTy improvements: **<span class="stat-inline">`r nrow(misty_imp_df)`</span>** rows
(HVG + B-cell runs).
Importances: **<span class="stat-inline">`r nrow(misty_importances)`</span>** rows
`r if(nrow(misty_importances)==0) '— mistyR not available or no importances collected.' else ''`.
</div>

```{r misty_display, results="asis"}
if (nrow(misty_imp_df) > 0) {
  for (run in unique(misty_imp_df$misty_run)) {
    sub <- misty_imp_df %>% filter(misty_run==run) %>%
      select(sample_id, treatment, timepoint, any_of(c("view","target","gain.R2","gain.RMSE","intra.R2","multi.R2"))) %>%
      arrange(sample_id, dplyr::desc(gain.R2 %||% 0))
    cat("\n\n### MISTy improvements —", run, "\n\n")
    print(kt(sub, caption=paste("MISTy view improvements —", run, "run")))
  }
}

if (nrow(misty_importances) > 0) {
  cat("\n\n### MISTy importances (paraview top predictors)\n\n")
  top_imp <- misty_importances %>%
    rename_with(tolower) %>%
    filter(grepl("para", view %||% "x", ignore.case=TRUE)) %>%
    group_by(sample_id, target %||% "target") %>%
    slice_max(abs(importance %||% 0), n=5, with_ties=FALSE) %>% ungroup() %>%
    arrange(sample_id, target)
  print(kt(head(top_imp, 80), caption="MISTy top 5 paraview predictors per target (head 80)"))
}
```

---

# 6. Manifest

```{r manifest}
files <- list.files(merged_tables_abs, pattern="\\.csv$", full.names=FALSE)
manifest <- tibble(
  file       = files,
  rows       = vapply(files, function(f) nrow(readr::read_csv(
    file.path(merged_tables_abs, f), show_col_types=FALSE, progress=FALSE)), integer(1)),
  size_kb    = round(file.size(file.path(merged_tables_abs, files)) / 1024, 1),
  written_at = format(file.mtime(file.path(merged_tables_abs, files)), "%Y-%m-%d %H:%M")
)
kt(manifest, caption=paste0("Merged_Tables/ — ", nrow(manifest), " files"))
```

```{r session}
writeLines(capture.output(sessionInfo()),
           file.path(merged_tables_abs, "session_info.txt"))
```
